\LARGE Untangling the link between traits, size and growth rate in plants

### 2016-11-14 ms/ms.tex
Incorporate feedback from Gibert

Plant species differ in many functional traits that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth and shade-tolerance, extending a widely-used theoretical model that links growth rate in seedlings with a single leaf trait to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Based on biomass production and allocation, this framework explains why the influence of prominent traits on growth rate and shade tolerance often varies with plant size and why the impact of size on growth varies among traits. Considering  growth rate in height, we find the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) wood density is present across sizes but is strongest at intermediate sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results. By disentangling the effects of plant size, light environment and traits on growth rates, our results provide a solid theoretical foundation for trait ecology and thus provide a platform for understanding growth across diverse species around the world.

@@

Functional traits capture core differences in the strategies plants use to generate and invest resources \citep{Westoby-2002, Wright-2004, Chave-2009}. Although most woody plants have the same basic physiological function and key resource requirements (carbon, nutrients and water), species differ considerably in the rates at which resources are acquired and invested into different tissues. During the last two decades, trade-offs related to some prominent traits have been quantified, with values for traits such as leaf mass per unit leaf area, wood density, and seed size now available for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. As data has accumulated, researchers are increasingly looking to traits to predict patterns in plant growth, demography, life history and performance \citep{Poorter-2008, Wright-2010, VanKleunen-2010, Adler-2014}. In this article we outline how the growth of individual plants is influenced by various functional traits, as well as plant size and the light environment.

@@

Recently, it has become clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations by \citet{Gibert-2016} provides the most compelling evidence. \citet{Gibert-2016} showed that the strength of the correlation between some traits (including LMA and maximum height) and growth rate was modified by size, while for other traits (including wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Importantly, the direction of the correlations and their shifts with size supported predictions generated via a framework describing how traits influence growth.

@@

One possible reason that theoretical predictions have either been lacking or not been supported is that the effects of traits in existing theory are realised mainly via influences on primary productivity (photosynthesis - respiration) \citep{Wright-2000, Enquist-2007}. By contrast, the physiology of traits such as LMA and wood density suggests they have influence by altering the allocation of biomass among different tissues, rather than changing mass production \citep{Falster-2011, Duursma-2016, Gibert-2016}. A second concern for theory focussed on mass production is that measuring mass production is really only practical for small plants that can be easily harvested. For larger plants, growth is more often measured as increment in either diameter or height. For example, stem diameter growth has been measured for many decades across many thousands of plots across the globe \citep{Purves-2008, Anderson-2015, Kunstler-2016}. Theory considering only primary production therefore offers limited insight.

Here we show how a new mechanistic growth model -- called {\plant} \citep{Falster-2016} -- can explain diverse empirical phenomena, including a size-dependent effect of traits on growth and an effect of traits on shade tolerance (Table \ref{tab:phenomena}), and thereby offer new insights into the way traits influence plant demography across the life-cycle. Broadly, the {\plant} model builds on several approaches to modelling production and allocation of biomass \citep[e.g.][]{Givnish-1988, Yokozawa-1995,Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011}. Our primary focus in this article is to explain the pattern that has been gradually emerging -- that the effect of traits on plant growth is modified by plant size \citep{Ruger-2012, Iida-2014, Gibert-2016}. Based on the same decomposition of growth rates as is implemented in the {\plant} model and used below \citep[from][]{Falster-2011}, \citet{Gibert-2016} argued conceptually why the effect of traits on growth should change with size. Here we extend the results of \citet{Gibert-2016} to provide a full, functioning model and use this to show, from the point of view of primary production and allocation, how and why the effect of some traits on growth rates changes with size. We also show how our approach can account for other phenomena, including changes in growth and shade tolerance with traits, individual size, and light environment. Our view is that trait-based approaches -- which aim to explain differences among species -- should be integrated within a general model of plant growth, and thus should also be able to capture patterns of growth through ontogeny. Growth rates tend to show hump-shaped relationship with size, when expressed as either height \citep{Sillett-2010, King-2011} or diameter growth \citep{Canham-2004, Canham-2006, Herault-2011} or biomass production \citep{Givnish-1988, Koch-2004}. In contrast, the growth rate of standing plant mass continues to increase with size \citep{Sillett-2010, Stephenson-2014}. All growth measures decrease sharply with size when expressed as relative growth rates \citep{Rees-2010, Iida-2014}. Shade tolerance also varies among species, correlates with traits \citep{Messier-1999, Lusk-2008, Poorter-2006}, and tends to decrease with increasing size \citep{Givnish-1988, Kneeshaw-2006, Lusk-2008}. These diverse phenomena -- summarised in Table \ref{tab:phenomena} -- are deserving of a comprehensive and integrated explanation.

@@

For the most part, parameters used in the current analysis were sourced from \citet{Falster-2016} (see Tables \ref{tab:params_core}--\ref{tab:params_hyper} for values). The only exceptions are: i) parameters affecting the relationships outlined in Table \ref{tab:allometry}a, which were estimated from data described below, and ii) parameters describing the function for reproductive allocation. By default, we set $r_{\rm r1}=0.8$ and $r_{\rm r2}=10$, implying a relatively rapid transition to reproduction at $H=H_{{\rm mat}}$ (see panel for $\frac{{\rm d}M_{{\rm a}}}{{\rm d}B}$ in Fig. \ref{fig:conceptual}).

The functional-balance assumptions listed in Table \ref{tab:allometry}a were evaluated using data from the Biomass and Allometry Database (\textsc{baad}) \citep{Falster-2015b}, which includes records for various size metrics from 21084 individual plants across 656 species. We fit standardised major axis ({\sma}) lines \citep{Warton-2006} to characterise bivariate relationships. We implemented a hierarchical model structure, where the distribution of slopes and intercepts among groups is assumed to come from normal distributions. The means and variances of these distributions are then fit as part of the model-fitting procedure.

@@

Our growth model suggests an intrinsically size-dependent pattern of biomass-production and growth, which aligns with well-known empirical patterns (Table \ref{tab:phenomena}). The panels in Fig. \ref{fig:conceptual} show the expected patterns for a typical woody plant, obtained by applying the functional-balance model encoded in {\plant}. Biomass production shows a hump-shaped pattern with size, decreasing at larger sizes as the turnover and respiration of sapwood and bark increase. Height growth also shows a hump-shaped pattern with size, first increasing then decreasing. This pattern results from systematic changes in the four components of eqn \ref{eq:dhdt} with size, including a strong decline in the fraction of plant that is leaf declines with increasing size (Fig. \ref{fig:conceptual}), increasing reproductive allocation (Fig. \ref{fig:conceptual}), and declining mass production. In contrast, basal-area growth continues to increase with size, due to an increasing influence of stem turnover. Diameter growth shows a weakly hump shaped curve, tapering off slightly at larger sizes, in part because of the allometric conversion from basal area to diameter (eqn \ref{eq:dDdt}, and in part because of increased reproductive allocation in older trees (Fig \ref{fig:conceptual}). All growth measures decrease sharply with size when expressed as relative growth rates (results not shown).

@@

Using a model relating plant physiological function and carbon allocation to five prominent traits, we have shown how traits impact on plant growth across the life cycle. This approach extends a widely-used theoretical model for seedlings, which links mass-based growth rate to the trait leaf mass per unit leaf area \citep{Lambers-1992, Wright-2000}, to explicitly include influences of size, light environment, and other prominent traits. During the last two decades, functional traits have captured the attention of ecologists, in large part because of the ability to organise the world's plant species along standard dimensions \citep{Westoby-2002}. However, it has remained unclear how or whether prominent traits influence growth outcomes \citep{Poorter-2008, Wright-2010,Paine-2015}. Matching a growing amount of empirical evidence (Table \ref{tab:phenomena}), this study outlines when and why the direction or strength of correlation between traits and growth rate shifts with plant size. Moreover, we show that different traits and trade-offs generate different types of response. Combined with the available empirical evidence, these results demand a fundamental shift in our understanding of plant ecological strategies, away from one in which species are thought to have a fixed growth strategy throughout their life (from slow to fast growth) \citep[e.g.][]{Grime-1977, Adler-2014, Paine-2015} to one in which traits define a size-dependent growth trajectory \citep{Gibert-2016}. Moreover, we find that growth trajectories and the ranking of traits across them are also moderated by the light environment; while traits that minimise costs of tissue respiration and / or turnover also make plants more shade tolerant (i.e. lower \textsc{wplcp}), as is empirically observed \citep{Messier-1999, Craine-2005, Poorter-2006, Baltzer-2007, Lusk-2008}. The {\plant} model, used here, builds on and extends several related approaches, wherein emergent outcomes such as height, diameter and mass growth arise from the interaction of different tissues and traits \citep[e.g.][]{Givnish-1988, Makela-1997, Moorcroft-2001}. This approach is quite different to models derived from metabolic scaling theory (\textsc{mst}), which derive everything from a single master ``scaling'' equation for mass growth and have thus far been unable to account for size-dependent changes in the correlation between traits and growth rate \citep{Enquist-1999, Enquist-2007}. Our approach is also fundamentally different from statistically-fitted growth models \citep[e.g.][]{Herault-2011, Ruger-2012, Iida-2014} in that it predicts rather than statistically tests for trait-based effects. In this sense, our model is designed to both explained observed phenomena (Table \ref{tab:phenomena}) and also generate new hypotheses.

>>>>

Plant species differ in many functional traits that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth and shade-tolerance, extending a widely-used theoretical model that links growth rate in seedlings with a single leaf trait to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Based on biomass production and allocation, this framework explains why the influence of prominent traits on growth rate and shade tolerance often varies with plant size and why the impact of size on growth varies among traits. Considering  growth rate in height, we demonstrate why the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) wood density is present across sizes but is strongest at intermediate sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results. By disentangling the effects of plant size, light environment and traits on growth rates, our results provide a solid theoretical foundation for understanding growth across diverse species around the world.

@@

Functional traits capture core differences in the strategies plants use to generate and invest resources \citep{Westoby-2002, Wright-2004, Chave-2009}. Although most woody plants have the same basic physiological function and key resource requirements (carbon, nutrients and water), species differ considerably in the rates at which resources are acquired, invested into different tissues, and lost via turnover. During the last two decades, trade-offs related to some prominent traits have been quantified, with values for traits such as leaf mass per unit leaf area, wood density, and seed size now available for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. As data has accumulated, researchers are increasingly looking to traits to predict patterns in plant growth, demography, life history and performance \citep{Poorter-2008, Wright-2010, VanKleunen-2010, Adler-2014}. In this article we outline the mechanisms by which the growth of individual plants is influenced by various functional traits, as well as plant size and the light environment.

@@

Recently, it has become clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations by \citet{Gibert-2016} provides the most compelling evidence. \citet{Gibert-2016} showed that the strength of the correlation between some traits (including LMA and maximum height) and growth rate was modified by size, while for other traits (including wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Importantly, \citet{Gibert-2016} proposed putative mechanism to explain their results. In this paper, we test these mechanisms by providing a fully functioning model of plant growth.

@@

One possible reason that theoretical predictions have either been lacking or not been supported is that the effects of traits in existing theory are realised mainly via influences on primary productivity (photosynthesis - respiration) \citep{Wright-2000, Enquist-2007}. By contrast, the physiology of traits such as LMA and wood density suggests they have influence by altering the allocation of biomass among different tissues, rather than changing mass production \citep{Falster-2011, Duursma-2016, Gibert-2016}. A second concern for theory focussed on mass production is that measuring mass production is really only practical for small plants that can be easily harvested. For larger plants, growth is more often measured as increment in either diameter or height. For example, stem diameter growth has been measured for many decades across many thousands of plots across the globe \citep{Purves-2008, Anderson-2015, Kunstler-2016}. As we outline below, mass production is only one of the factors influencing diameter growth; theory considering only primary production therefore offers limited insight.

Here we show how a new mechanistic growth model -- called {\plant} \citep{Falster-2016} -- can explain diverse empirical phenomena, including a size-dependent effect of traits on growth and an effect of traits on shade tolerance (Table \ref{tab:phenomena}), and thereby offer new insights into the way traits influence plant demography across the life-cycle. Broadly, the {\plant} model builds on several approaches to modelling production and allocation of biomass \citep[e.g.][]{Givnish-1988, Yokozawa-1995,Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011, Gibert-2016}. Our primary focus in this article is to explain the pattern that has been gradually emerging -- that the effect of traits on plant growth is modified by plant size \citep{Ruger-2012, Iida-2014, Gibert-2016}. Based on the same decomposition of growth rates as is implemented in the {\plant} model and used below \citep[from][]{Falster-2011}, \citet{Gibert-2016} argued conceptually why the effect of traits on growth should change with size. Here we extend the results of \citet{Gibert-2016} to provide a full, functioning model and use this to show, from the point of view of primary production and allocation, how and why the effect of some traits on growth rates changes with size. We also show how our approach can account for other phenomena, including changes in growth and shade tolerance with traits, individual size, and light environment. Our view is that trait-based approaches -- which aim to explain differences among species -- should be integrated within a general model of plant growth, and thus should also be able to capture patterns of growth through ontogeny. Growth rates tend to show hump-shaped relationship with size, when expressed as either height \citep{Sillett-2010, King-2011} or diameter growth \citep{Canham-2004, Canham-2006, Herault-2011} or biomass production \citep{Givnish-1988, Koch-2004}. In contrast, the growth rate of standing plant mass continues to increase with size \citep{Sillett-2010, Stephenson-2014}. All growth measures decrease sharply with size when expressed as relative growth rates \citep{Rees-2010, Iida-2014}. Shade tolerance also varies among species, correlates with traits \citep{Messier-1999, Lusk-2008, Poorter-2006}, and tends to decrease with increasing size \citep{Givnish-1988, Kneeshaw-2006, Lusk-2008}. These diverse phenomena -- summarised in Table \ref{tab:phenomena} -- are deserving of a comprehensive and integrated explanation.

@@

For the most part, parameters used in the current analysis were sourced from \citet{Falster-2016} (see Tables \ref{tab:params_core}--\ref{tab:params_hyper} for values). The only exceptions are: i) parameters affecting the relationships between various size metrics and leaf area, outlined in Table \ref{tab:allometry}a, estimated from data described below, and ii) parameters describing the function for reproductive allocation. By default, we set $r_{\rm r1}=0.8$ and $r_{\rm r2}=10$, implying a relatively rapid transition to reproduction at $H=H_{{\rm mat}}$ (see panel for $\frac{{\rm d}M_{{\rm a}}}{{\rm d}B}$ in Fig. \ref{fig:conceptual}).

The functional-balance assumptions listed in Table \ref{tab:allometry}a were evaluated using data from the Biomass and Allometry Database \citep[\textsc{baad}][]{Falster-2015b}, which includes records for various size metrics from 21084 individual plants across 656 species. We fit standardised major axis ({\sma}) lines \citep{Warton-2006} to characterise bivariate relationships. We implemented a hierarchical model structure, where the distribution of slopes and intercepts among groups is assumed to come from normal distributions. The means and variances of these distributions are then fit as part of the model-fitting procedure.

@@

Our growth model suggests an intrinsically size-dependent pattern of biomass-production and growth, which aligns with well-known empirical patterns (Table \ref{tab:phenomena}). The panels in Fig. \ref{fig:conceptual} show the expected patterns for a typical woody plant, obtained by applying the functional-balance model encoded in {\plant}. Biomass production shows a hump-shaped pattern with size, decreasing at larger sizes as the turnover and respiration of sapwood and bark increase. Height growth also shows a hump-shaped pattern with size, first increasing then decreasing. This pattern results from systematic changes in the four components of eqn \ref{eq:dhdt} with size, including a strong decline in the fraction of plant that is leaf declines with increasing size (Fig. \ref{fig:conceptual}), increasing reproductive allocation (Fig. \ref{fig:conceptual}), and declining mass production. In contrast, basal-area growth continues to increase with size, due to an increasing influence of stem turnover. Diameter growth shows a weakly hump shaped curve, tapering off slightly at larger sizes, in part because of the allometric conversion from basal area to diameter (eqn \ref{eq:dDdt}), and in part because of increased reproductive allocation in older trees (Fig \ref{fig:conceptual}). All growth measures decrease sharply with size when expressed as relative growth rates (results not shown).

@@

Using a model relating plant physiological function and carbon allocation to five prominent traits, we have shown how traits impact on plant growth across the life cycle. This approach extends a widely-used theoretical model for seedlings, which links mass-based growth rate to the trait leaf mass per unit leaf area \citep{Lambers-1992, Wright-2000}, to explicitly include influences of size, light environment, and other prominent traits. During the last two decades, functional traits have captured the attention of ecologists, in large part because of the ability to organise the world's plant species along standard dimensions \citep{Westoby-2002}. However, it has remained unclear how or whether prominent traits influence growth outcomes \citep{Poorter-2008, Wright-2010,Paine-2015}. Matching a growing amount of empirical evidence (Table \ref{tab:phenomena}), this study outlines when and why the direction or strength of correlation between traits and growth rate shifts with plant size. Moreover, we show that different traits and trade-offs generate different types of response. Combined with the available empirical evidence, these results demand a fundamental shift in our understanding of plant ecological strategies, away from one in which species are thought to have a fixed growth strategy throughout their life (from slow to fast growth) \citep[e.g.][]{Grime-1977, Adler-2014, Paine-2015} to one in which traits define a size-dependent growth trajectory \citep{Gibert-2016}. Moreover, we find that growth trajectories and the ranking of traits across them are also moderated by the light environment; while traits that minimise costs of tissue respiration and / or turnover also make plants more shade tolerant (i.e. lower \textsc{wplcp}), as is empirically observed \citep{Messier-1999, Craine-2005, Poorter-2006, Baltzer-2007, Lusk-2008}. The {\plant} model, used here, builds on and extends several related approaches, wherein emergent outcomes such as height, diameter and mass growth arise from the interaction of different tissues and traits \citep[e.g.][]{Givnish-1988, Makela-1997, Moorcroft-2001}. This approach is quite different to models derived from metabolic scaling theory (\textsc{mst}), which derive everything from a single master ``scaling'' equation for mass growth and have thus far been unable to account for size-dependent changes in the correlation between traits and growth rate \citep{Enquist-1999, Enquist-2007}. Our approach is also fundamentally different from statistically-fitted growth models \citep[e.g.][]{Herault-2011, Ruger-2012, Iida-2014} in that it predicts rather than statistically tests for trait-based effects. In this sense, our model is designed to both explain observed phenomena (Table \ref{tab:phenomena}) and also generate new hypotheses.
